---
title: "Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant"
author:
  - Eric R. Scott:
      email: scottericr@gmail.com
      institute: [UF]
      correspondence: true
  - Paulo Rubim:
      email: rubimpaulo@gmail.com
      institute: [UFSC]
      correspondence: false
  - Maria Uriarte:
      email: mu2126@columbia.edu
      institute: [Columbia]
      correspondence: false
  - Emilio M. Bruna:
      email: embruna@ufl.edu
      intitute: [UF]
      correspondence: false
  
institute:
  - UF: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - UFSC: Departamento do Botânica, Universidade Federal de São Carlos, CP 676, CEP 13565-905, São Carlos, SP, Brazil
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/ecology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
highlights: |
  These are the highlights
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

## Habitat fragmentation

Habitat fragmentation is bad, pervasive, and getting worse

-   stats about global land use change <!--# start this broad or start with the Amazon? -->

-   stats about amazon deforestation [@malhi2008]

-   stats about how much land is near an edge (refs in Schwartz et al and betts 2019)

-   Why fragmentation bad

    -   microclimate changes [@ewersFragmentationImpairsMicroclimate2013]

    -   metapopulation and connectivity stuff (necessary?)

-   Plants in the tropics are less well adapted to fragmentation because lack of historical distrubance and less variable climate [@betts2019]

## Climate change and drought

-   The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].

-   The El Niño--Southern Oscillation (ENSO) reduces wet-season rainfall in the Amazon <!--# is this correct? Or only in the south? -->. ENSO-induced droughts are predicted to increase in frequency and severity. [@caiIncreasingFrequencyExtreme2014].

-   Also sea surface temperature related drought. Warming of north Atlantic results longer and more intense dry season in southern and eastern Amazonia [@phillipsDroughtSensitivityAmazon2009; @malhi2008]

-   drought effects on plant vital rates

-   1997 ENSO drought [@williamsonAmazonianTreeMortality2000]

-   2005 ENSO drought [@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008]

-   2010 drought [@lewis2010AmazonDrought2011]

-   [@zengCausesImpacts20052008]

## Interactions

-   However, our understanding of how climate change and habitat fragmentation might interact to impact organisms is limited.

-   [@schwartz2019] (mention how tree size classes affected differently, but we don't know if this is also true for understory herbaceous plants.

-   [@malhi2008]

-   The 1997--1998 ENSO drought caused greater tree mortality and increased leaf litter fall in experimentally fragmented forests compared to intact forest in the Amazon [@laurancePositiveFeedbacksForest2001].

-   Furthermore, the interactive effects of fragmentation and drought may be difficult to detect in long-lived species (e.g. Heliconia) [bruna paper that shows growth is reduced after 10 years slightly, but impacts reproduction]

## This experiment

-   The "fragmentation drought hypothesis" expects drought to have stronger impacts on forest fragments since fragments less buffered from changes in humidity and more sensitive to desiccation. Under this hypothesis we would expect drought to reduce growth, survival, reproduction and recruitment more in fragmented landscapes compared to intact forest.

-   It's also possible that plants in fragments have already been filtered such that the survivors are already adapted to more variable and desiccating conditions associated with forest edges [@schwartz2019]. This may result in drought having a less noticeable impact on plants in fragments, although we would expect fragmentation to still have an impact on survival, growth, and reproduction.

-   We set out to answer the following questions: 1) Does drought have different impacts on plants in forest fragments compared to intact forest? 2) Do the magnitudes and directions of these effects differ among vital rates? 3) Do the effects of drought, fragmentation, and their interaction vary with plant size [@tredennickSizebyenvironmentInteractionsNeglected2018] <!--# read this referrebce --> ?

# Background

# Methods

*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible understory monocot native to the nonflooded rainforests of central Amazonia. *Heliconia* are rhizomatious, but clonal reproduction via underground runners is rare in *H. acuminata*, unlike other congeners [@bruna2004]. Vegetative shoots grow from the rhizome as well as one or more flowering shoots when plants reproduce. Each flowering shoot has a single inflorescence of 20--25 flowers [@bruna2002]. Flowers are pollinated nearly exclusively by two species of hummingbirds at this site <!--# species? -->. Pollinated ovaries develop into fleshy fruits containing a maximum of 3 seeds which are dispersed exclusively by birds [@uriarte2011; @brunaHeliconiaAcuminataSeed2014].

## Site

-   All field data were collected at the Biological Dynamics of Forest Fragments Project (BDFFP)---a long term experimental site located about 80km north of Manaus (2º30' S, 60ºW) [@laurance2011; @laurance2002; @bierregaard1992]. The experimental plots are located in BDFFP reserves in nonflooded tropical lowland rainforest with nutrient-poor, well-draining, xanthic ferrosols.

-   Something about seasonality

-   BDFFP is comprised of replicated forest fragments of 1, 10, and 100ha as well as continuous forest plots that were established between 1980 and 1984. Fragments were maintained by clearing secondary growth around fragments occasionally <!--# Best reference for this? -->. Demographic plots ($50 \times 100$m) were established in <!--# year --> to track the growth, reproduction, and survival of *Heliconia acuminata*. Six of these plots were established in continuous forest, four were in 1-ha fragments, and three were in 10-ha fragments. The plots in continuous forest were established at haphazardly chosen locations 500--4000m from any mature forest borders, plots in 1-ha fragments were established on a randomly selected half of the fragment, and plots in 10-ha fragments were established in the center.

# Results

# Discussion

# Conclusion

# Acknowledgments

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
