---
title: "Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant"
author:
  - Eric R. Scott:
      email: scottericr@gmail.com
      institute: [UF]
      correspondence: true
  - Maria Uriarte:
      email: mu2126@columbia.edu
      institute: [Columbia]
      correspondence: false
  - Emilio M. Bruna:
      email: embruna@ufl.edu
      intitute: [UF]
      correspondence: false
  
institute:
  - UF: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
linkcolor: blue
bibliography: references.bib
csl: "../templates/ecology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
highlights: |
  These are the highlights
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

1.  Organisms experience multiple sources of natural and anthropogenic stress simultaneously. However, the impacts of multiple sources of stress on population dynamics are not well understood [@tyeDemographicMenageTrois2016].

    -   Cite Pedro F. Quintana-Ascencio in intro---he will likely be a reviewer.

    -   Cite Jennifer Williams

    -   Emphasize demography. We know about interactions on physiology and other stuff, but we don't know as much about how demography of plants is shaped by how populations respond to multiple stresses, increasingly anthropogenic.

    -   Especially difficult when lagged effects are expected. Part of the challenge in doing this modeling is the statistics are new <!--# maybe goes in discussion --> . "Move beyond the statistical shackles of interaction terms".LOL

2.  As a result of intense deforestation, tropical lowland forests are becoming an increasingly fragmented landscape. This is especially true in the Amazon basin where \_\_\_\_ km^2^ forest are within 100m of a forest edge <!--# citation -->.

    -   "Over 70000 km of new forest edges are being created annually in the Brazilian Amazon by human activities (Broadbent et al., 2008)"

    -   also cite: Kapos 1989, Lovejoy et al. 1986, Kapos et al. 1997?

    -   This can result in ...

3.  Simultaneously, these fragmented habitats are experiencing increased drought stress as a result of climate change. The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008]. Additionally, El Niño Southern Oscillation (ENSO)-induced droughts are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].

    -   1997 ENSO drought [@williamsonAmazonianTreeMortality2000]

    -   2005 ENSO drought [@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008]

    -   2010 drought [@lewis2010AmazonDrought2011]

    -   Drought is bad because...

    -   Combined environmental stresses, such as drought and fragmentation, have the potential to exacerbate or dampen effects on population dynamics [@tyeDemographicMenageTrois2016].

4.  Plant populations may be more susceptible to drought in fragmented forests. Fragmentation results in changes in forest microclimate that could exacerbate the effects of drought on plant populations. Near edges, tropical forests experience increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].

    -   Trees near forest edges experience increased mortality after droughts compared to forest interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].

    -   Increased mortality after droughts disproportionately affects large trees.

5.  Alternatively, plant individuals that are more susceptible to desiccation may not have survived fragmentation, resulting in populations in fragments which are more resilient to drought.

    -   [@betts2019].

6.  <!--# I'm now thinking this doesn't need to be a separate paragraph, but included in #4.  Need to do some more reading about this though. -->

    In addition to variation among vital rates, responses to drought and fragmentation could depend on life history stage or plant size. Large established adults could respond differently than seedlings. This is unknown because there are very few systems where we have data on entire life history in a climate change or fragmentation context.

    -   Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]). Smaller/larger plants might be more susceptible to drought. So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.

7.  We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape. Specifically, we want to know 1) Does drought increase or decrease the growth, survival, and fertility rates of plant populations in continuous forest? 2) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest? 3) Are the effects of drought and fragmentation on vital rates consistent across life-history stages?

# Methods

<!--# Emilio, feel free to edit -->

*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible understory monocot native to the nonflooded rainforests of central Amazonia. *Heliconia* are rhizomatious, but clonal reproduction via underground runners is rare in *H. acuminata*, unlike other congeners [@bruna2004]. Vegetative shoots grow from the rhizome as well as one or more flowering shoots when plants reproduce. Each flowering shoot has a single inflorescence of 20--25 flowers [@bruna2002]. Flowers are pollinated nearly exclusively by two species of hummingbirds at this site <!--# species? -->. Pollinated ovaries develop into fleshy fruits containing a maximum of 3 seeds which are dispersed exclusively by birds [@uriarte2011; @brunaHeliconiaAcuminataSeed2014].

## Site

<!--# Emilio: please edit -->

-   All field data were collected at the Biological Dynamics of Forest Fragments Project (BDFFP)---a long term experimental site located about 80km north of Manaus (2º30' S, 60ºW) [@laurance2011; @laurance2002; @bierregaard1992]. The experimental plots are located in BDFFP reserves in nonflooded tropical lowland rainforest with nutrient-poor, well-draining, xanthic ferrosols.

-   Something about seasonality

-   BDFFP is comprised of replicated forest fragments of 1, 10, and 100ha as well as continuous forest plots that were established between 1980 and 1984. Fragments were maintained by clearing secondary growth around fragments occasionally <!--# Best reference for this? -->. Demographic plots ($50 \times 100$m) were established in <!--# year --> to track the growth, reproduction, and survival of *Heliconia acuminata*. Six of these plots were established in continuous forest, four were in 1-ha fragments, and three were in 10-ha fragments. The plots in continuous forest were established at haphazardly chosen locations 500--4000m from any mature forest borders, plots in 1-ha fragments were established on a randomly selected half of the fragment, and plots in 10-ha fragments were established in the center.

# Results

# Discussion

metapopulation dynamics stuff goes here.

# Conclusion

# Acknowledgments

If TRMM data is used, they suggest: "The TMPA data were provided by the\
NASA/Goddard Space Flight Center's Mesoscale Atmospheric Processes Laboratory and\
PPS, which develop and compute the TMPA as a contribution to TRMM."

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
