---
title: "Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape"
author:
  - Eric R. Scott:
      email: scottericr@gmail.com
      institute: [UF_WEC]
      correspondence: true
  - María Uriarte:
      email: mu2126@columbia.edu
      institute: [Columbia]
      correspondence: false
  - Emilio M. Bruna:
      email: embruna@ufl.edu
      institute: [UF_WEC, UF_LAS, BDFFP]
      correspondence: false
  
institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
linkcolor: blue
bibliography: references.bib
csl: "templates/ecology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
highlights: |
  These are the highlights

editor_options:
  markdown:
    wrap: sentence
    canonical: true
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>",
	dpi = 300
)
library(targets)
library(knitr)
library(here)
```

```{r load-targets}
withr::with_dir(here(), {
  tar_load(c(
    normals_plot,
    eda_plot,
    s_spei_plot,
    s_spei_diff_plot,
    g_spei_plot,
    g_spei_diff_plot,
    f_spei_plot,
    f_spei_diff_plot,
    size_plot,
    f_cf,
    f_1ha
  ))
})
```

<!-- The actual document text starts here: -->

<!--# Consider co-authorship of Cris Jurinitz, Maria Rosa Darrigo, Paulo Rubim, John Kress. -->

# Introduction

Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015].
Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
Although the local extinction of species from individual fragments is common [@haddadHabitatFragmentationIts2015]<!--# others? -->, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009]<!--# others? -->.
However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments -- e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] -- can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others? -->.
This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
Despite this, evidence linking changes population-level demographic responses with changes in environmental conditions remains scarce.

<!--# CITATION_4: WRI Report? Or INPE, see here: https://rainforests.mongabay.com/amazon/deforestation_calculations.html or this: https://mapbiomas.org/en -->

Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome. For instance, <!--#parts of th      e Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50-100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013]. These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, which are unable to move in response to changes in local conditions. In contrast, other parts of the basin are predicted to see an increase in the frequency and severity of extreme precipitation events, which can also have detrimental consequences for plant survival and reproduction [@estebanOtherSideDroughts2021].

Add here what is known about frag and drought effects - without setting it up so people expect a pval for an intearaciton tern

Whether plant populations in tropical forest fragments are particularly susceptible to drought remains unclear for three primary reasons.
First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
These data are critical not simply because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @tellerLinkingDemographyDrivers2016].
They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [@gagnonGrowthUnderstoryHerb2011].
Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021; @andereggPervasiveDroughtLegacies2015; @schwalmGlobalPatternsDrought2017; @kannenbergGhostsHowDrought2020].

Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***] We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
This time series, which included the severe droughts of 1997 [@mcphadenChildProdigy1997981999] and 2005 [@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008], allowed us to address the following questions:\

1)  Does drought increase or decrease the growth, survival, and fertility of plant populations in continuous forest?

2)  Are there delayed effects of drought on demographic vital rates, and if so what lag times are most critical?

3)  Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?

# Methods

## Study site

The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
In addition to large areas of continuous forest, the BDFFP has forest fragment reserves isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried (Figure S\_ <!--# map can go in the supplemental materials -->).
In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].

```{r}
cap <- "Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981–2010. Precipitation (a) shows a marked dry seasons from June through October. Temperature (b) varies less throughout the year. Mean monthly temperature is shown in the solid red line and monthly minimum and maximum temperatures are shown with the lower and upper dashed lines, respectively. Data from Brazilian National Institute of Meterology (INMET)."
```

```{r normals, fig.cap=cap}
normals_plot
```

The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy [@rankin-de-meronaPreliminaryResultsLargescale1992] and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content.
Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm.
There is a pronounced dry season from June to October (Figure \@ref(fig:normals)).

## Focal species

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others---such as *H. acuminata*---are found at lower densities in the darker and cooler forest understory.
These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991].
While some *Heliconia* species are at least partially self-compatible [@kressSelfincompatibilitySystemsCentral1983], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare *[unpubl. data]*.

In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
Dispersed seeds germinate \~6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].

## Demographic data collection

In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
This study uses data from 1-ha fragments and Continuous Forest sites.

Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and the identity of any flowering plants.
We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002].
Plants that were not found for three consecutive surveys were considered dead.
There is a seven-fold range in *H. acuminata* density in the forest plots, with density in several of the continuous forest plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].

## Climate data

Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmmTMPA3B432011]), though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
<!--# maybe add a sentence emphasizing that SPEI is monthly calculation -->

## Statistical Modeling of Vital Rates

<!-- I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).  -->

To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
Using the *dlnm* package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
We chose 36 months as a maximum lag because prior transplant experiments with *H. acuminata* showed they typically recovered from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.

<!-- re 36 mos transpalnt shock justification; how are lag lengths typically chosen?  -->

The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the *mgcv* package [@woodGeneralizedAdditiveModels2017].
The general form of the vital rate models was as follows:

$$
\begin{split}
& g[ \textrm{E}(y_{i})] = \alpha_j + s_1(x_i) + s_2(d_{i,t}, ..., d_{i,t-36}) \\
& \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \;\textrm{for plot }j 
\end{split}
$$

where $s_1(x_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al. 2017 for details).
The crossbasis function, $s_2(\cdot)$ can also be written:

$$
s_2(d_{i,t}, ..., d_{i,t-24}) = \sum_{l = l_0}^L f \cdot w(d_{i,t-l},l) 
$$

where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see Wood [-@woodChapterSmoothers] and Gasparrini et al. [-@gasparriniPenalizedFrameworkDistributed2017] for more detail).
Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e. number of months) and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal when most complex <!--# You may want to say what the 3 knots can capture: increasing then decreasing, exp decline etc -->.
Because of penalization, the number of knots is generally not important as long it is great enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].

To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$.
We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
This was modeled with a scaled t family error distribution and the natural log of height $\times$ shoot number as our proxy for plant size.
A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
For reproduction in year t our response variable was whether or not a plant had any inflorescences (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function .

In the process of fitting the models, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity [@woodGeneralizedAdditiveModels2017].
We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in Marra and Wood [-@marraPracticalVariableSelection2011].

The `dlnm` package does not currently allow modeling of factor by smooth interactions.
This means we cannot \_\_\_\_\_\_\_\_\_\_\_\_\_\_.
We therefore fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.

# Results

```{r}
cap <- "Timeseries of demographic parameters and drought ocurrence. Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (red) than continuous forest (blue). (b) shows the survival curve for the 1998 cohort of plants in both fragments and continuous forest. c) shows the proportion of plants with 4 or more shoots flowering each census. In d), monthly 3-month SPEI is plotted. Light grey lines represent values from different grid cells encompassing BDFFP and the dark line is the site mean. Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
```

```{r eda, fig.cap=cap, fig.height=7, fig.width=10}
eda_plot
```

<!--# consider boxplots or median ± IQR for panel a) -->

Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature.
For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\
The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).
<!--#  I know I'm missing some literature here. Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->

## Survival, growth, and flowering in continuous forest vs. fragments

**Our dataset comprised XXXX plants in continuous forest and XXXX plants in forest fragments, with an plots in CF having on average X-fold more plants than plots in 1-ha fragments (mean per plot + SD in each) .** When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
The proportion of surviving plants was lowest in the 2003-2004 transition year (P(surv) = 0.92), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d).
The lowest survival for 1-ha fragment plots (p(surv) = 0.93) was for the 2005-2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in DOUBLE CHJECK DATES OF DROUGHT AND RAIN. Survival was size dependent in both continuous forest and 1-ha fragments (p \< 0.001 for the log-transformed effect of plant size in year t on suvival in year t+1).
While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.

<!--# The proportion of surviving plants was lowest from 2004-2005 (0.92), which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan–Mar of 2004 (these dates overlap? -->

<!--# This is across the whole timeseries, which I'm not sure is actually that important. What summary statistics would be useful? -->

Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average **\~XX% fewer** shoots (2.5 ± 1.5 SD) and **were XX% shorter** (36.2 cm ± 24.1 SD).
Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently significantly larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively).
This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).

<!--# Is growth really higher in fragments?  Probably, since fig2 shows size becoming more similar over time.   -->

While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (0.050 ± 0.021 vs. 0.036 ± 0.027, respectively).
This was true in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold value of log 5\_\_\_\_ (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots.
Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest vs. only 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
The back-transformed model intercepts, representing the flowering probability of averaged sized plants in each habitat given average SPEI, showed a similar pattern with 0.008 for continuous forest and 0.004 for 1-ha fragments.
<!--# not sure how necessary this is. EB: same-->

<!-- This paragraphis talkng about size, but not really growth -->

<!--# There was a pronounced threshold in flowering probability corresponding with a log size of about 5...is this 5 shoots? What abvout the proxy size? -->

## Delayed effects of drought on demographic vital rates

```{r}
cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probabiility, modeled by a cubic regression splines. Panel d) shows the smoothed density of different plant sizes in the two habitat types. 95% confidence intervals are shown and include uncertainty in the model intercept and uncertainty due to smoothness selection. The smooths for 1-ha fragments and continuous forest are fit in separate models. The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale.  In b), a 1:1 line is shown in dashed red to indicate the line of no growth."
```

```{r size-plot, fig.cap=cap, fig.height=8, fig.width=5}
size_plot
```

```{r}
cap <- "Smooth effect of lagged SPEI on survival in a) continuous forest and b) 1-ha fragments. Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis. Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale. Contour lines correspond to a change of 0.002. The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."

```

```{r surv-spei, fig.cap=cap, fig.height=5, fig.width=7}
s_spei_plot
```

Drought history had a significant effect on survival in both habitats (P \< 0.001), but the shape of the crossbasis surface differed.
For 1-ha fragments, recent SPEI values (i.e. within the past 0--12 months) had the strongest effect on survival, with the highest survival near SPEI of 0 and decreasing survival when plants were exposed to either drier or wetter conditions over the prior 0--12 months (Figure \@ref(fig:surv-spei)a).
In continuous forest, recent SPEI had a weaker effect with only SPEI values less than -1 leading to slightly reduced survival probabilities (Figure \@ref(fig:surv-spei)b).
However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year (i.e. not the most recent dry season, but the one before that), also affected survival probabilities in continuous forest.
Lower SPEI (drought) 15--20 months prior to census was associated with reduced survival and higher SPEI at these lag times was associated with higher survival.
Plants in both continuous forest and 1-ha fragments showed an increase in survival probability with very high (wet) SPEI values at a lag time of 36 months.
However, it should be noted that only the first year of census data (1999) had an SPEI value greater than 2 within 36 months in the past.

```{r}
cap <- "The difference between predicted effects of SPEI history on survival in continuous forest and 1-ha fragments."
```

```{r surv-diff, fig.cap=cap, fig.height=2.5, fig.width=7}
s_spei_diff_plot
```

Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-diff)).
However, under extreme conditions, survival probability is higher in continuous forest by up to 0.02.

```{r}
cap <- "Smooth effect of lagged SPEI on plant growth for a) continuous forest and b) 1-ha fragments. Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis. Model intercepts were added to fitted values of the crossbasis function. Contour lines correspond to a change of 0.05. The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
```

```{r growth-spei, fig.cap=cap, fig.height=5, fig.width=7}
g_spei_plot
```

Drought history also had a significant effect on plant size in both habitats (p \< 0.001).
The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 5.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
Drought at lags 8--11 (March - June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.

```{r}
cap <- "The difference between predicted effects of SPEI history on plant size in continuous forest and 1-ha fragments."
```

```{r growth-diff, fig.cap=cap, fig.height=2.5, fig.width=7}
g_spei_diff_plot
```

The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-diff).
However, growth in continuous forests is higher by up to log(size) = 0.57 when the current wet season is unusually wet (SPEI\>2).
This is due to reduced growth in fragments under wet conditions.

```{r}
cap <- "Smooth effect of lagged SPEI on flowering probability. Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis. Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale. Contour lines correspond to a change of 0.001. The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
```

```{r flwr-spei, fig.cap=cap, fig.height=5, fig.width=7}
f_spei_plot
```

Drought history had a significant effect on flowering probability in both habitat types (p \< 0.001).
Flowering probability was overall higher at all values of SPEI in continuous forest compared to 1-ha fragments (Figure \@ref(fig:flwr-spei)).
The effects of SPEI on flowering probability were also stronger and more complex in continuous forest compared to fragments (1-ha edf = 8.5, continuous forest edf = 10.5).
In continuous forest, flowering probability was increased by recent drought (at lag = 0 with SPEI \< -1) and drought at lags 15--20 (Jun-Nov, two dry seasons ago).
Drought at lags 7--13 reduced flowering probability.
The shape of the crossbasis function was similar in 1-ha fragments, but significantly muted (Figure \@ref(fig:flwr-spei)a).
That is, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.

```{r}
cap <- "The difference between predicted effects of SPEI history on flowering probability in continuous forest and 1-ha fragments."
```

```{r flwr-diff, fig.cap=cap, fig.height=2.5, fig.width=7}
f_spei_diff_plot
```

Plotting the difference between the crossbasis smooths for continuous forest and 1-ha fragments revealed that flowering probability is higher in continuous forest compared to 1-ha fragments and is increased by recent drought and drought during the dry season 15--20 months prior to the census (Figure \@ref(fig:flwr-diff)).
The least difference between habitat types occurs when there is strong drought at about 8--12 months lag or very wet conditions at 15--20 months lag.

FInally, with the exception of growth in 1-ha fragments (p = 0.257), the delayed effects of SPEI on all three vital rates varied significantly between plots (p \< 0.001 for the random effect of plot).

# Discussion

Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought (-1.5 \> SPEI \> -2) during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
Two noticeable peaks in the proportion of flowering mature plants occur in severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).

#### Vocab

This seems tacked on here and perhaps better in the discussion: populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].

\< !--
\# Gaoue = ?
Orou G. Gaoue, Carol C. Horvitz, Ulrich K. Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,<https://doi.org/10.1016/j.ecolmodel.2019.03.007>.
-- \>

Delayed effects, time lag, the lag...avoid lagged effects

#### Citations / information to allocate to appropriate paragraphs

-   @criley1994 : under hort conditions, shoots that flower in spring emerge in June/July the previous year.
    \~6 months after shoot emergence that the developing infloresence is susceptible to adverse effects.

-   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod. can be affected

-   Lots of great stuff about the importance of rhizomes, including for drought resistance, in @klimesovaBelowgroundPlantFunctional2018.
    Rhizome resistance to variability in environmental conditions also a conclusion in this study on invasion by another zingiberales ( @pinheiroEffectRhizomeExposure2021

-   @westerbandEarlyLifeConditions2017: *Heliconia* grew more at low than high precipitation.
    Leaf life span and biomass allocation did not differ among treatments.
    The effect of environmental variability was dependent on early life conditions as well as precipitation, suggesting that generalist species may experience high fitness as forest environments become more variable by maintaining high growth at the expense of survival.

-   the observed effects on reproduction can have effects beyond demography, eg.
    gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007)

-   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975

-   @wrightSeasonalDroughtSoil1992: seasonal drought.
    experimental irrigation influences survival and growth of tropical herbs

-   Effects of precip extremes on understory plants can be also be indirect:

    -   high precip = lots of clouds, less photosynthesis, less or slower growth

    -   low precip = trees lose leaves or die, more light gets through or gaps open.
        This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses2020)

-   @horvitzDemographicCostReproduction1988: no evidence for short term (next season) cost of reproduction in another understory monocot (Calathea) BUT

-   @horvitzEffectsPlantSize2002 (*Calathea ovandensis*) Survival of juveniles was positively affected by plant size, but not by herbivory, local competition, or fruit production.
    Survival of large stages was not generally affected by any of the factors.
    Relative growth rates of all stages were negatively affected by plant size; larger plants grew more slowly than smaller plants.
    Additionally, growth rate of juveniles was negatively affected by local competition, whereas growth rate of reproductives was positively affected by reproduction.
    Next year's inflorescence production was positively affected by plant size for both pre-reproductive and reproductive stages.
    For reproductives, it was additionally positively affected by current fruit production.

-   @rundelStructuralPhysiologicalAdaptation1998 Heliconia appear to allocate much of their above-ground biomass to leaf tissues and have a relatively low investment in support tissues.
    Investment in leaf form vs structure interpreted as a trade-off between economy and protection against tissue loss from falling branches; presence of below-ground rhizomes in Heliconia may be key.
    See also @rundelFunctionalTraitsBroadleaved2020\

-   @stiles1975?
    drought resulted in less flowering in Heliconia

-   @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet

-   @wrightSeasonalNinoLonger2006: Increase in flower and seed production after El Niño events for trees and lianas.
    Might be due to increases solar irradiance.
    Maybe evidence of lagged effects?
    Table 2 is inconsistent with text, so not sure...

-   Many humid tropical forest species time reproduction to coincide with dry season peak to avoid low irradiance in wet season.
    (refs within @wrightSeasonalNinoLonger2006. van Shaik et al. 1993; Wright and vanSchaik 1994).
    But *Heliconia* times reproduction so seeds only germinate at start of rainy season.

-   [@estebanOtherSideDroughts2021] negative effects of large positive SPEI

-   In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.

-   refs in @leitoldNinoDroughtIncreased2018 peak in tree mortality can happen up to 3 years after drought event in experimental studies.
    Observational studies show mortality happening in same year as 2007 and 2010 droughts.

-   @leitoldNinoDroughtIncreased2018 increased canopy turnover during drought.
    Similar increase in continuous and (large) fragments.
    Increase in turnover more variable in fragments.
    Coarse woody debris increased 62% during drought.\

<!--# in intro, questions are 1: what are effects on vital rates, 2: are effects delayed, 3: differences between habitats.  Keep the same order here? -->

Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013]. <!--# more here -->. Despite this, surprisingly few studies have compared the structure and demography of populations in habitat fragments and continuous habitat; those that have are typically short term (i.e., \<4 years). Using a decadal dataset of plant demography, we show that some - but not all - results from short-term studies can be extrapolated to longer times scales. Our results also add to the emerging body of work showing delayed effects of climate drivers on demographic vital rates [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Finally, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.

Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
Using a decadal dataset of plant demography, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.

# Temporal variation in demographic responses to forest fragmentation

Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002]. In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002]. While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher. The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences. However, while in our initial three-year survey the relationship between plant size and probability of flowering was similar in both habitat types [@brunaHabitatFragmentationDemographic2002], when using the decadal dataset were found that flowering probability for the largest plants is actually higher in forest fragments (i.e., average fitted probability of flowering for plants in upper size quartile = 0.125 in CF vs. 0.094 in FF). These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance - populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002]. However, these differences in flowering probability could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest. Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations [@gagnonGrowthUnderstoryHerb2011].

# Delayed effects of climate on demographic vital rates

Precipitation and other climatic factors are known to have large effects on plant growth, survival, or reproduction.
Despite this, models of population dynamics have rarely included the effects of environmental drivers.
This has largely due to the challenge - both ecologically and statistically - of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021]. Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature -- that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.

However, while it appeared that delayed effects of climate on demographic vital rates were ubiquitous, the extent to which they vary spatially or with habitat remained an open question [@eversLaggedDormantSeason2021]. Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates. For example, [delete this phrase? the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI - both positive (unusually high precipitation) and negative (drought conditions) - led to declines in the probability of individual survival both habitat types.
However, the magnitude of these declines was far greater in forest fragments.
Similarly, the detrimental effects on growth rates of variation in SPEI, in this case unusually rainy wet seasons and dry seasons with unusually low precipitation, were also more severe in fragments.
In contrast, variation in SPEI had a far stronger effect on populations in continuous forests, though it is is important to recognize that because there are fewer plants in fragments above the threshold-size for flowering it is possible we lacked the statistical power to detect the full effects of SPEI in fragments.

Both variation in precipitation and extreme conditions are predicted to increase as climate change progresses ( <!--# cite IPCC? -->) and our results demonstrate that this may have a greater negative impact on understory plants in forest fragments compared to continuous forests.

```{r flwr-prob-quartile, include=FALSE}

#figure out flowering probability of largest quartile of plants in each habitat
library(tidyverse)
newdf_cf <-
  f_cf$model %>% #use model data frame
  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
  filter(quartiles == 4)

newdf_1ha <-
  f_1ha$model %>% 
  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
  filter(quartiles == 4)

#equivalent to flowering once every __ years
1/predict(f_cf, newdata = newdf_cf) %>% mean() %>% plogis()
1/predict(f_1ha, newdata = newdf_1ha) %>% mean() %>% plogis()
```

Reduced microclimate buffering could possibly explain the greater effects of variation in SPEI in fragments compared to continuous forest.
A major consequence of forest fragmentation is reported alteration in microclimate near forest edges [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013]. It is possible that increases in temperature, evaporative drying, wind speed, and solar radiation commonly found near forest edges exacerbated the effects of extremes in SPEI on *H. acuminata* growth and survival in fragments compared to continuous forest. It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1 ha fragments. Bruna et al. [-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H. accuminata* plants transplanted near edges was reduced relative to continuous forest.
A greater understanding of the microclimatic changes in fragments during extreme climate events and their effects on understory plants is needed, but we believe differences in microclimate buffering are a likely explanation for the observed differences between fragments and continuous forest.

Past climate conditions were important for survival, size, and flowering probability.
Extremes in SPEI in the 15 months prior to the census reduced survival in forest fragments (Figure \@ref(fig:surv-spei)).
Extremes in SPEI in the past 36 months affected plant size, although the direction of the effects was seasonally dependent (Figure \@ref(fig:growth-spei)).
Generally, unusually dry dry seasons and unusually wet wet seasons had negative impacts on plant size, and this pattern was more extreme in fragments compared to continuous forest.
For flowering, the SPEI of the current wet season and the dry season two dry seasons ago (lag of 15--20 months) had the strongest influence on flowering probability with dry weather corresponding to an increase in flowering (Figure \@ref(fig:flwr-spei))<!--# wondering if this is more of a results paragraph? Alternatively it goes *after* the next paragraph and includes more speculation about why each specific delayed effect exists.  Might end up integrating with a later paragraph. -->.

There are several possible non-mutually-exclusive explanations for delayed effects of SPEI on vital rates.
First, delayed effects may represent a demographic delay since we do not know precisely when since the previous survey the plant died or changed size.
For example, if extreme drought in the dry season before the census increases plant mortality in that season, this would show up as a delayed effect in our models (e.g. in Figure \@ref(fig:surv-spei)b) despite the impact of drought potentially being immediate.
This explanation for delayed effects applies only to plant size and survival, as flowering occurs very close to the yearly demographic census (Jan--Apr).

Second, these delayed effects may represent a physiological delayed effect of environmental conditions.
Under horticultural conditions, flowering shoots of *Heliconia chartacea* that flower in Spring begin to emerge in June or July of the previous year, and the developing inflorescence may be aborted in response to adverse conditions up to 6 months after initiation [@crileyYearProductionHigh1994]. <!--# more here about general trends in physiological delays for plants witht tubers or rhizomes -->Although flowering occurs in late January through April in *H. acuminata*, the initiation of a primordial flowering shoot <!--# not sure this is the right term.  "Differentiation of the apical meristem" ? --> likely happens many months earlier. The increase in flowering rate of plants following a drought could also be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018] <!--# add another more general citation. e.g. S. Joe Wright trees flowering after ENSO -->.
We also considered that some delayed effects could be a result of a cost of reproduction leading to reduced growth, survival, and/or flowering probability in the following season.
However, when flowering in the previous season is included as a covariate, it predicts higher growth and flowering probability---likely a reflection of size being a good predictor of these vital rates (Supplemental\_\_\_).
Our models show that drought two dry seasons prior to the census increase flowering probability.
A greater understanding of the effects of light, drought, season, and their interactions on flowering in *H. acuminata* is needed to provide a probable mechanism for this pattern.

Finally, delayed effects could be mediated by indirect effects and ecological interactions.
<!--# I'm not sure if this is really an explanation for the *delay* per se.  Need to think about this. -->Drought in tropical forests can increase tree mortality and leaf drop, resulting in decreased canopy cover and increased light penetration to the understory (<!--# citations -->).
This increase in light could then result in increased growth or flowering for understory plants ( <!--# citations -->).
<!--# Heliconia grow better at edges, where there's more light --> Janssen et al. [@janssenDroughtEffectsLeaf2021] showed that increase in litterfall due to drought was delayed.
Here, the delayed effect of drought on *H. acuminata* could be partly explained by a delay of the response of canopy trees to drought and the response of *H. acuminata* to increased insolation.
Furthermore, literfall in response to drought has shown to be greater at forest edges compared to interiors, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments.
Drought is also associated with increased insolation, so negative SPEI could also be a proxy for increased insolation, which would be exacerbated at forest edges potentially.

-   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets.
    Estimates green-up, literfall, growth, etc. for central and southern america including the Amazon.
    Drought results in anomolously high leaf flushing at start of drought followed by increased litterfall durring the drought.
    Dry season droughts result in green-up, but not wet season droughts.
    During drought, old leaves are shed and new leaves are maintained.
    2-5 months for newly flushed leaves to fully expand.

-   Ecological Consequences of these patterns: an eye towards a drier/wetter future

# Conclusions & Future Directions

-   There could be size-specific effects: Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartzFragmentationForestStructure2019]).
    Smaller/larger plants might be more susceptible to drought.
    So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
    The `dlnm` package does not currently allow for modeling of interactions, and we lacked data on plant size at the same time scale as SPEI (monthly), so interactions with size were not included in this analysis.

-   Synergistic effects of fragmentation and climate (need the stats to catch up with us) "Organisms simultaneously experience multiple forms of disturbance. While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood. They could act in concert to exacerbate or dampen population dynamics."

-    few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
    This demonstrates the importance of long term demographic experiments in determining the effects of changing habitats on populations.
    <!--# Or something like "some of these long-term studies exist and are prime for doing these kinds of analyses" -->

-   

-   Loop back to: need whole-life-cycle approach to ID these effects, why fragmentologists don't do this is killing Emilio.

# Acknowledgments

We thank ----, ----, ----, and --- anonymous reviewers for helpful discussions and comments on the manuscript.
We also thank the technicians and students who helped conduct the *Heliconia* censuses and the BDFFP and INPA for logistical support.
Financial support was provided by the U.S.
National Science Foundation (awards DEB-0614339, DEB-0614149, INT 98-06351, and DEB-1948607).
This is publication No. ------ in the BDFFP Technical Series.

If TRMM data is used, they suggest: "The TMPA data were provided by the\
NASA/Goddard Space Flight Center's Mesoscale Atmospheric Processes Laboratory and\
PPS, which develop and compute the TMPA as a contribution to TRMM.
"

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
