---
title: "Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments"
author:
  - Eric R. Scott:
      email: scottericr@gmail.com
      institute: [UF_WEC]
      correspondence: true
  - María Uriarte:
      email: mu2126@columbia.edu
      institute: [Columbia]
      correspondence: false
  - Emilio M. Bruna:
      email: embruna@ufl.edu
      institute: [UF_WEC, UF_LAS, BDFFP]
      correspondence: false
institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      linenumbers: yes
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
      - -Fpandoc-crossref
linkcolor: blue
bibliography: references.bib
csl: "templates/global-change-biology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract. 34,281 year x individual measurements! 
keywords: |
  Amazon; delayed effects; demography; distributed lag non-linear models; drought; ENSO; habitat fragmentation; *Heliconia acuminata*; Heliconiaceae
highlights: |
  These are the highlights
  
editor_options:
  markdown:
    wrap: sentence
    canonical: true
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

<!-- Highlights: `r rmarkdown::metadata$highlights` -->

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>",
	dpi = 300
)
library(targets)
library(tidyverse)
library(tsibble)
library(lubridate)
library(knitr)
library(here)
library(patchwork)
```

```{r load-targets}
withr::with_dir(here::here(), {
  tar_load(c(
    eda_plot,
    surv_curve,
    s_spei_plot,
    g_spei_plot,
    f_spei_plot,
    size_plot,
    f_cf,
    f_1ha,
    model_data,
    xa_spei
  ))
})
```

<!-- The actual document text starts here: -->

<!--# Consider co-authorship of Cris Jurinitz, Maria Rosa Darrigo, Paulo Rubim, John Kress. -->

\pagebreak

# Introduction

Tropical deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015; @alroy2017].
Deforestation also results in landscapes where the remaining forest can be highly fragmented, with forest remnants of different sizes embedded in a matrix of often highly-contrasting habitat [@taubert2018; @bianchi2013].
For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
Fragmentation is associated with myriad ecological changes, including the local and regional extinction of plant species [@laurance2006; @dasilva2000].
Although the demographic mechanisms responsible these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009], it is often hypothesized that the dramatically altered environmental conditions in fragments [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] are driving declines in reproduction, recruitment, or survivorship [@brunaSeedGerminationRainforest1999 ; @zartmanDroughtImpactsSurvivorship2015; @lauranceRainForestFragmentation1998].
Despite the fact that this mechanism is thought to be particularly important in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999; @bettsExtinctionFiltersMediate2019], efforts to link population-level demographic responses with altered environmental conditions remains scarce.

Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwood2015; @williams2015], with more pronounced effects for fragmented populations [@oliver2015; @holyoak2016].
While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015].
Plant populations may be particularly sensitive -- an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
This, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004; @selwood2015].

Whether the demography of plant populations in tropical forest fragments is more susceptible to drought and other climatic extremes remains unclear for three primary reasons.
First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
Similarly, there is a growing literature on how tropical plants respond to droughts [@esquivel-muelbertCompositionalResponseAmazon2019; @gonzalez-mDivergingFunctionalStrategies2020; @uriarteImpactsClimateVariability2016], but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
These data are critical not simply because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @tellerLinkingDemographyDrivers2016].
They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [*e.g.,* @gagnonGrowthUnderstoryHerb2011].
Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021; @andereggPervasiveDroughtLegacies2015; @schwalmGlobalPatternsDrought2017; @kannenbergGhostsHowDrought2020].

Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [@brunaHabitatFragmentationPlant2009].
We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
This time series, which included the severe droughts of 1997 [@mcphadenChildProdigy1997981999] and 2005 [@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008], allowed us to address the following questions: (1) Does drought increase or decrease the growth, survival, and fertility of plant populations in continuous forest?
(2) Are there delayed effects of drought on demographic vital rates, and if so what lag times are most critical?
(3) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?

# Methods

## Study site

The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
In addition to large areas of continuous forest, the BDFFP has forest fragment reserves isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried [@bierregaardBiologicalDynamicsTropical1992].
In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].

The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy [@rankin-de-meronaPreliminaryResultsLargescale1992] and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content.
Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall ranges from 1900--2300 mm.
There is a pronounced dry season from June to October (Figure S1 <!--# update figure number manually -->).

## Focal species

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
While many *Heliconia* species grow in large patches in treefall gaps and other disturbed areas, others, such as *H. acuminata,* are found at lower densities in the darker and cooler forest understory [@rundelFunctionalTraitsBroadleaved2020].
These species produce fewer inflorescences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996; @brunaHeliconiaAcuminataReproductive2004].

In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}=1.1$ flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}=2$), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
Dispersed seeds germinate \~6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].

## Demographic data collection

In 1997--1998 we established a series of 5000 m^2^ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
The plots are located in continuous forest ($n=6$), 10 ha fragments ($n=3$), and 1-ha fragments ($n=4$), with distance between plots ranging from 500 m--41 km.
Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10 ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500--4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
This study uses data from 1-ha fragments and continuous forest sites.
<!--# maria: justify why 10ha plots omitted or don't say they exist -->

Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]).
In this study, we used the product of shoot number and plant height as our measure of plant size.
Preliminary analysis showed that the product of shoot number and height was a better predictor of total leaf area (which in turn is assumed to be a strong predictor of aboveground biomass) than either shoot number or height alone (Table S2 <!--# update crossref manually -->).
Plants that were not found for three consecutive surveys were considered dead.
We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
For additional details on the plots, survey methods, and *H. acuminata* populations see @brunaHabitatFragmentationDemographic2002.

## Climate data

Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º $\times$ 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
Note that 3-month SPEI is still monthly data---each month's SPEI value simply takes into account precipitation and evapotranspiration of the previous three months.
SPEI calculations were made using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought can sometimes differ (Figure S2; Table S1 <!--# update crossrefs manually -->).

## Statistical Modeling of Vital Rates

To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
Using the `dlnm` package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
We chose 36 months as a maximum lag because prior transplant experiments with *H. acuminata* showed they typically recover from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.

The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
The general form of the vital rate ($y$) models was as follows:

$$
  \begin{split}
  & g[ \textrm{E}(y_{i})] = \alpha_j + s_1(z_i) + s_2(d_{i,t}, ..., d_{i,t-36}) + \beta x_i \\
  & \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \;\textrm{for plot }j 
  \end{split}
$${\#eq:eqn1}

where $s_1(z_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al. 2017 for details).
The crossbasis function, $s_2(\cdot)$ can also be written:

$$
s_2(d_{i,t}, ..., d_{i,t-24}) = \sum_{l = l_0}^L f \cdot w(d_{i,t-l},l) 
$${\#eq:eqn2}

where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see chapter 5 of @woodGeneralizedAdditiveModels2017 and @gasparriniPenalizedFrameworkDistributed2017 for more detail).
Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e. number of lagged SPEI values for each observation with 36 months as a maximum lag) and 3 knots for the drought response dimension to restrict the shape of the fitted response to drought to bimodal when most complex.
Because of penalization, the number of knots is generally not important as long it is large enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
Estimated degrees of freedom (edf) represent the 'true' complexity of the smooth after penalization with edf = 1 being equivalent to a straight line and larger numbers representing more complex curves.

To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$ in @eq:eqn1.
We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
This was modeled with a scaled t family error distribution because residuals were leptokurtic with a Gaussian error structure.
Because number of inflorescences was highly zero-inflated, we converted this to a binary response to model reproduction (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function.
We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.

In the process of fitting the models, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity [@woodGeneralizedAdditiveModels2017].
We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in @marraPracticalVariableSelection2011.

The `dlnm` package does not currently allow modeling of factor by smooth interactions.
This means we could not include habitat as an interaction term.
We therefore fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.

# Results

Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature (Table S1).
For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).
The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).

## Survival, growth, and flowering in continuous forest vs. fragments

```{r include=FALSE}
summary_stats <-
  model_data %>%
  group_by(habitat) %>%
  summarize(across(c(ht, shts, size, flwr, surv),
                   .fns = list(mean = mean, sd = sd),
                   na.rm = TRUE),
            n_plants = length(unique(ha_id_number))) %>%
  # right_join(model_data %>% count(habitat)) %>%
  mutate(across(ends_with("_mean") | ends_with("_sd"),
                                ~round(.x, 2)))
summary_stats <- 
  full_join(
  summary_stats,
  model_data %>%
    count(habitat, plot) %>%
    group_by(habitat) %>%
    summarize(n_mean = mean(n), n_sd = sd(n)) %>%
    mutate(across(where(is.numeric), round))
)
summary_stats

by_year <-
  model_data %>% 
  group_by(year, habitat) %>% 
  summarize(prop_surv = mean(surv)) %>% 
    arrange(prop_surv)
by_year

xa_spei %>% ungroup() %>%
  summarize(across(everything(), mean)) %>%
  filter(year(yearmonth)==2005)

```

Our dataset comprised `r format(summary_stats$n_plants[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n_plants[1], big.mark = ",")` plants in forest fragments.
Plots in CF had on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` SD; 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` SD).

When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high ($P_{surv} = 0.95$; Figure \@ref(fig:eda)b).
The proportion of surviving plants was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on survival in year t+1 in both habitats).
While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)b), the survival of the smallest plants was higher in 1-ha fragments.
However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.

Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average \~`r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were \~`r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
Because the proxy for plant size used in our models was the product of these two metrics, plants in continuous forest were consequently substantially larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively).
<!--# Perhaps worth mentioning here that their distribution in different size classes is also different (skewed smaller) and likely why they are smaller on average). -->This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage.
Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$ in both habitats).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)a).

```{r include=FALSE}
#data for eda plot panel c
model_data %>% 
  filter(size >= 168) %>% 
    group_by(year, habitat) %>% 
    summarize(n = n(), flwr = sum(flwr == 1, na.rm = TRUE)) %>% 
    mutate(prop_flwr = flwr/n)
```

While the overall proportion of all plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (log(size) \> 5 in Figure \@ref(fig:size-plot)c).
Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs. only 13.6% in 1-ha fragments (Figure \@ref(fig:eda)c).

## Delayed effects of drought on demographic vital rates

Drought history had a significant effect on the survival ($p < 0.001$), growth ($p < 0.001$), and flowering ($p < 0.001$) of plants in both habitats.
Comparing the respective crossbasis surfaces, however, reveals that the specific climatic drivers, their timing, and their impact on individual vital rates all differed among habitats.

<!--# what do you think of this wording? Trying to walk people through the big result by breaking into two sentences telling them what they are about to see -->

For 1-ha fragments, recent SPEI values (i.e. within the past 0--12 months) had the strongest effect on survival <!--# Is another way to say this "For 1-ha fragments, SPEI in the preceding 12 months had the strongest effect on survival, with..."  -->., with the highest survival near SPEI of 0 and mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)a).
<!--# what do you think? trying to put the results in more ecological terms) -->In contrast, the effect of recent SPEI in continuous forest was both weaker and unidirectional -- survival probability was reduced only when SPEI was \< -1 (i.e., mild-severe drought), and even then only slightly (Figure \@ref(fig:surv-spei)b).
Instead the most pronounced effects on the survival of plants in continuous forest were of SPEI at a lag time of 15--20 months (i.e. two dry seasons prior to a census).
Drought 15--20 months prior to a census (i.e., SPEI \< -1) was associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was associated with higher survival (put another way, was a nearly linear relationship between SPEI and survival probability).
Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 36 months.
It should be noted, however, that only the first year of census data (1999) met these conditions.
<!--# can you walk me through these next two sentences? I'm not sure if this applies to the 36 mop lag or the whole time-series -->Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-spei)c).
However, under extreme conditions, survival probability is higher in continuous forest by up to 0.02.

While the fitted crossbasis functions for fragments and continuous forest showed generally similar patterns of drought effects on growth (i.e, trends in plant size), the surface for 1-ha fragments indicated more complex responses in some situations (edf = 17.9 for 1-ha fragments; edf = 12.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
For example, when SPEI = 0 (i.e., average conditions), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
However, when the current wet season is unusually wet (SPEI\>2), plants in continuous forests were larger by up to log(size) = 0.57 due to reduced growth in fragments.
In contrast, drought at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats, but the magnitude of the response was greater in 1-ha fragments.

Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)).
The effects of SPEI on flowering probability in continuous forest were also more complex (1-ha edf = 8.1, continuous forest edf = 10.3); recent drought (i.e., at lag = 0 with SPEI \< -1) increased the probability of flowering in continuous forest, as did drought at lags 15--20 (i.e., two dry seasons prior).
However, drought at lags 7--13 <!--# this corresponds to what - the dry and rainy 1 year ago?  --> reduced flowering probability.
The responses in 1-ha fragments were substantially more muted, as indicated by the shape of the crossbasis function (Figure \@ref(fig:flwr-spei)a).
In other words, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.
Plotting the difference between the crossbasis smooths for continuous forest and 1-ha fragments revealed that flowering probability is higher in continuous forest compared to 1-ha fragments and is increased by recent drought and drought during the dry season 15--20 months prior to the census (Figure \@ref(fig:flwr-spei)c).
The least difference between habitat types occurs when there is strong drought at about 8--12 months lag or very wet conditions at 15--20 months lag.

***ALTERNATIVE SUMMARY OF THE FLOWERING RESULTS?*** Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)).
The responses in 1-ha fragments were also more muted (1-ha edf = 8.1, continuous forest edf = 10.3), as indicated by the shape of the crossbasis function (Figure \@ref(fig:flwr-spei)a).
This led to some important inter-habitat differences in plant responses to prior droughts: recent drought (i.e., at lag = 0 with SPEI \< -1) increased the probability of flowering in continuous forest, as did drought at lags 15--20 (i.e., two dry seasons prior).
However, drought at lags 7--13 <!--# this corresponds to what - the dry and rainy 1 year ago?  --> reduced flowering probability in continuous forest far more than it did in fragments (Figure \@ref(fig:flwr-spei)c).
The greatest similarity between forest and fragments occurs when there is strong drought 8--12 months prior, or precipitation 15--20 months prior is extremely high.

FInally, with the exception of survival in 1-ha fragments ($p = 0.252$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.001$ for the random effect of plot).
There was no evidence for a cost of reproduction --- in both forest and fragments, flowering in the previous year was a significant predictor of growth (CF: $p = 0.048$, 1-ha: $p = 0.031$) and flowering ($p < 0.001$ for both CF and 1-ha).
However, the coefficients were always positive, indicating that flowering in the previous year was positively related to both size and flowering probability in the current year.
This suggests that....<!--# not surew what to make of this lastsentence. Interpret for reader?   -->

# Discussion

Understanding how landscape structure and abiotic conditions act to influence population dynamics is central to many conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
Our results support the emerging consensus that the effects of climatic extremes on demographic vital rates can be delayed for months or even years [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
We also found that the delayed responses of populations in fragments can differ significantly in magnitude, direction, and lag time from those of populations in continuous forest.
This suggests that the hypothesized synergies between climate and fragmentation on population dynamics [@opdam2004a; @selwood2015a; @laurance2001] are likely to be pervasive, but also far more complex than previously thought.

## Temporal variation in demographic responses to forest fragmentation

Our results indicate that some---but not all---results from short-term demographic studies can be extrapolated to longer times scales.
Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002].
In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002].
While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher.
The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences.
However, while in our initial three-year survey the relationship between plant size and probability of flowering was similar in both habitat types [@brunaHabitatFragmentationDemographic2002], when using the decadal dataset were found that flowering probability for the largest plants is actually higher in forest fragments (i.e., average fitted probability of flowering for plants in upper size quartile = 0.125 in CF vs. 0.094 in FF).

These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance---populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants.
They translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest (assuming stasis in size and average SPEI).
While some *Heliconia* species could compensate for these disparities with by selfing [@kressSelfincompatibilitySystemsCentral1983] or vegetative reproduction @rundelFunctionalTraitsBroadleaved2020], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare and this species does not reproduce clonally (Bruna *unpubl. data*).
Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations.
[@gagnonGrowthUnderstoryHerb2011].

<!--# new paragraph below.  What do you think Emilio? -->

Fragmentation of forests diminishes their ability to buffer changes in microclimate due to edge effects [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
This can result in changes in vegetation structure, and is a plausible mechanism for reduced growth and survival of *H. acuminata* in forest fragments in this study.
We show that the effects of variation in SPEI on survival and growth are more extreme in fragments compared to continuous forest.
Additionally, the direction of the SPEI effects appear to be seasonal---that is, extreme dry conditions in dry seasons and extreme wet conditions in wet seasons are most detrimental to growth and survival.
The ability of continuous forest habitat to buffer these extremes may be responsible for the less extreme effects of variation in SPEI in continuous forest.
More experimental work is needed to establish alterations in microclimate or impaired microclimate buffering in extreme weather events as a causal mechanism for reduced growth and survival of plant individuals in forest fragments.

## Delayed effects of climate on demographic vital rates

Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
These effects can be complex or even contradictory---mild droughts can increase the growth rates of tropical trees and seedling survival, perhaps due to reductions in cloud cover and concomitant increases in solar radiation[@conditTropicalForestDynamics2004; @uriarteEnvironmentalHeterogeneityBiotic2018; @hueteAmazonRainforestsGreenup2006; @alfaro-sanchezGrowthReproductionRespond2017; @jonesAsynchronousAmazonForest2014], but in severe drought years growth can be extremely low and mortality can be sharply elevated [@connell2000; @engelbrecht2002; @edwards2006].
There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" [@wright1999; @pau2013].

Despite this knowledge, models of plant population dynamics have rarely included the effects of environmental drivers.
This has largely been due to the challenge (both ecologically and statistically) of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
To address this challenge, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to 3 growing seasons.

While it appears that delayed effects of climate on demographic vital rates may be ubiquitous [@eversLaggedDormantSeason2021], the extent to which they vary spatially or with habitat remains an open question.
Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
For example, although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival in both habitat types, the magnitude of these declines was far greater in forest fragments.
Similarly, the detrimental effects of extremes in SPEI on growth rates were also more pronounced in fragments.
In contrast, the effect of variation in SPEI on flowering probability was far stronger in continuous forest than fragments; these results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.

There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@eversLaggedDormantSeason2021], and hence be shaped by climatic events at any point in that physiological window.
For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 months prior to the appearance of inflorescences.
Adverse conditions during the 6 months following initiation, rather than the months when inflorescences are starting expand, leads to the aborted production of flowering shoots [@crileyYearProductionHigh1994].
Interestingly, we observed the opposite effect---drought conditions increased the probability of flowering two years later.
While this could reflect bet-hedging in response to stress [@nihadRelationshipStressFlowering2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988]. In fact, flowering in one year is associated with increased reproduction and growth in the next.

Demographic responses will also be delayed if abiotic stress causes plants to invest in below-ground rhizomes [*sensu* @pumisutapon2012].
The carbohydrates stored in rhizomes allow *Heliconia* to regenerate above-ground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimesovaBelowgroundPlantFunctional2018].
This may be why drought led to delayed increases in growth---by shedding shoots and leaves [@brunaExperimentalAssessmentHeliconia2002] and investing in rhizomes, plants are generating proportionately more buds with which to regenerate when conditions improve.
This would also be consistent with the results of prior experiments, in which the growth rates of *H. acuminata* 8 months after they were mechanically damaged far exceeded those of control plants [@brunaCompensatoryResponsesUnderstory2005].

Third, it may be that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on other species rather than the direct effects on individual physiology [@eversLaggedDormantSeason2021].
For example, while droughts can kill some tropical trees immediately [@lauranceEffectsStrongDrought2001], others take three years or more [@crileyYearProductionHigh1994].
This will potentially lead to a boom in growth and flowering years after the drought, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990; @leitoldNinoDroughtIncreased2018].
Similar delayed changes in the local environment could also influence the foraging behavior of *H. acuminata*'s pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996] and seed dispersers [@uriarteDisentanglingDriversReduced2011].
While more work is needed to explain why the (delayed) effects of SPEI on survival and growth are greater in forest fragments than interiors, it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.

Finally, demographic delays could be an artifact of the timing of responses in relation to the census date.
If extreme drought in the dry season before the census increased plant mortality during that season, for example, this would nevertheless appear in models as a delayed effect (e.g. in Figure \@ref(fig:surv-spei)b).
In our case, this potential explanation for delayed effects applies only to plant size and survival, as plots were surveyed regularly throughout the reproductive season to identify flowering plants.
This is possibility is not unique to our study, rather it is a consequence of conducting demographic censuses on an annual scale while the climate is quantified monthly or seasonally.
While the very slow growth and extremely low mortality rates of *H. acuminata* mean this effect is unlikely to be acting in our system, it may be that for some species it will be important to conduct demographic surveys at the same temporal scale at which climate is aggregated.

```{r flwr-prob-quartile, include=FALSE}

#figure out flowering probability of largest quartile of plants in each habitat
library(tidyverse)
newdf_cf <-
  f_cf$model %>% #use model data frame
  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
  filter(quartiles == 4)

newdf_1ha <-
  f_1ha$model %>% 
  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
  filter(quartiles == 4)

#equivalent to flowering once every __ years
1/predict(f_cf, newdata = newdf_cf) %>% mean() %>% plogis()
1/predict(f_1ha, newdata = newdf_1ha) %>% mean() %>% plogis()
```

## Conclusions & Future Directions

Climate models for the Amazon predict a future of extremes---increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015].
Our results support the hypothesis that populations in fragments could be more susceptible to the effects of changing climate than those in continuous forest [@lauranceApparentEnvironmentalSynergism2014].
However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.
Multi-factorial, multi-season experiments [*sensu* @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandEarlyLifeConditions2017; @aguirre2021], ideally manipulating multiple combinations of climatic variables [@mundim2016], are needed to determine how and why habitat-specific differences in environmental conditions interact to delay the demographic responses of plants to climatic variability.
Also needed are statistical tools that can test for synergistic effects of fragmentation and climate in vital rates, as those currently available do not allow for including interaction terms.
This also limits the ability to include size by climate interactions in a DLNM; although plant responses to both fragmentation and climatic extremes can be size-specific [@brunaDemographicEffectsHabitat2005; @schwartzFragmentationForestStructure2019].
The ability to identify size-specific lagged responses may be especially complicated given size and growth are rarely measured at the same time scale as SPEI and other putative climatic drivers.

Finally, no analytical approach assessing the potential for demographic lags can compensate for a lack of long-term data [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021].
Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.

# Acknowledgments

We thank Collin Edwards, ----, ----, and --- anonymous reviewers for helpful discussions and comments on the manuscript.
We also thank the technicians and students who helped conduct the *Heliconia* censuses and the BDFFP and INPA for logistical support.
Financial support was provided by the U.S.
National Science Foundation (awards DEB-0614339, DEB-0614149, INT 98-06351, and DEB-1948607).
This is publication No. ------ in the BDFFP Technical Series.

# Figures

```{r}
eda_cap <- "Timeseries of demographic parameters and drought occurrence. Mean fold-change in plant plant size (log2(size_t+1_ / size_t)) (a) shows that plant growth varied by year and habitat (1 ha fragments in orange, solid; continuous forest in blue, dashed). In most years plants grew on average with a notable exception in 2003 where plants regressed in size on average in both habitats (fold-change < 0). Error bars in a) represent standard deviation. The proportion of plants surviving each transition year (b) shows that average survival is high and was lowest in the 2004 census. The proportion of reproductive sized plants (c) is on average low and fluctuates substantially year to year. The cutoff for reproductive size plants in panel (c) is defined as the upper 90th percentile size of flowering plants in all years. Monthly 3-month SPEI is shown (d) with gray lines representing values from different grid cells encompassing BDFFP and the dark line representing the site mean. Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
```

```{r eda, fig.cap=eda_cap, fig.width=7, fig.height=7}
eda_plot & theme(text = element_text(size = 10))
```

```{r}
surv_curve_cap <- "Survivorship of plants labeled in the first survey year, 1998, which comprise 49% of the plants in the full dataset. After 10 years, 79.7% (1629/2055) of plants in continuous forest survived and 72.4% (393/543) of plants in fragments survived."
```

```{r surv_curve, fig.cap=surv_curve_cap, fig.width=7, fig.height=3.5}
surv_curve + theme(text = element_text(size = 10))
```

```{r}
size_cap <- "Smooth effect of plant size in the previous census from models for a) survival, b) log(size), and c) flowering probability, corresponding to the additive term s_1_(z_i_) in eq. 1.  95% confidence intervals are shown and include uncertainty in the intercept and uncertainty due to smoothness selection. The smooths for 1-ha fragments and continuous forest are fit in separate models. Panel d) shows the distribution of plant sizes in the two habitat types."
```

```{r size-plot, fig.cap=size_cap, fig.width=7, fig.height=7}
size_plot & theme(text = element_text(size = 10))
```

```{r}
s_spei_cap <- "Smooth effect of lagged SPEI on survival in a) continuous forest and b) 1-ha fragments. Panel c) shows the difference between panels a) and b). Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis. Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale. Contour lines correspond to a change of 0.002. The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."

```

```{r surv-spei, fig.cap=s_spei_cap, fig.height=7.5, fig.width=7}
s_spei_plot & theme(text = element_text(size = 10))
```

```{r}
g_spei_cap <- "Smooth effect of lagged SPEI on plant growth for a) continuous forest and b) 1-ha fragments. Panel c) shows the difference between panels a) and b). Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis. Model intercepts were added to fitted values of the crossbasis function. Contour lines correspond to a change of 0.05. The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
```

```{r growth-spei, fig.cap=g_spei_cap, fig.height=7.5, fig.width=7}
g_spei_plot & theme(text = element_text(size = 10))
```

```{r}
f_spei_cap <- "Smooth effect of lagged SPEI on flowering probability for a) continuous forest and b) 1-ha fragments. Panel c) shows the difference between panels a) and b).Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis. Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale. Contour lines correspond to a change of 0.001. The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
```

```{r flwr-spei, fig.cap=f_spei_cap, fig.height=7.5, fig.width=7}
f_spei_plot & theme(text = element_text(size = 10))
```

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

# Left-over text

-   populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].
-   While the alteration of microclimate near forest edges is one of the fundamental results to have emerged from the study of fragmented landscapes [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013], little is known regarding how variable it is. It We hypothesize that the microclimate in continuous forest plots is more stable than that in 1 ha fragments, and that this stability will demographically buffers populations in forest interiors [@brunaExperimentalAssessmentHeliconia2002].
-   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975
-   eg. gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007)
-   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets. Estimates green-up, literfall, growth, etc. for central and southern America including the Amazon. Drought results in anomalously high leaf flushing at start of drought followed by increased litterfall during the drought. Dry season droughts result in green-up, but not wet season droughts. During drought, old leaves are shed and new leaves are maintained. 2-5 months for newly flushed leaves to fully expand

# Citations / information could allocate to appropriate paragraphs

-   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentation influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod. can be affected

-   Effects of precip extremes on understory plants can be also be indirect:

    -   high precip = lots of clouds, less photosynthesis, less or slower growth

    -   low precip = trees lose leaves or die, more light gets through or gaps open.
        This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses2020

-   @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet

-   @wrightSeasonalNinoLonger2006: Increase in flower and seed production after El Niño events for trees and lianas.
    Might be due to increases solar irradiance.
    Maybe evidence of lagged effects?
    Table 2 is inconsistent with text, so not sure...

-   [@estebanOtherSideDroughts2021] negative effects of large positive SPEI

-   In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.

-   @leitoldNinoDroughtIncreased2018 increased canopy turnover during drought.
    Similar increase in continuous and (large) fragments.
    Increase in turnover more variable in fragments.
    Coarse woody debris increased 62% during drought.

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
