---
title: "Size-environment interaction"
author: "Eric R. Scott"
date: "2020-01-11"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
bibliography: "`r here::here('analysis', 'paper', 'references.bib')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(lme4)
library(bbmle)
library(car)
library(conflicted)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# TODO

- ~~Test that random effect of year is significant for all models~~

# Purpose

Rather than explicitly modeling drought effects on *Heliconia*, model environmental stochasticity as a random effect of year. This allows me to test an interaction term with size to see if responses to environment (maybe drought) are size dependent. Inspired by Tredennick et al. [-@tredennickSizebyenvironmentInteractionsNeglected2018].

# Load Data

This is the same demography data used elsewhere, but without the SPEI matrix-column

```{r data, echo=TRUE}
demog <-
  read_csv(
    here("analysis", "data", "derived_data", "ha_survey.csv"),
    col_names = TRUE,
    cols(plot = col_character(),
         bdffp_reserve_no = col_character())
  )
demog <-
  demog %>%
  filter(ht < 200, ht_next < 200) %>% 
  filter(shts > 0, shts_next > 0, ht > 0, ht_next > 0) %>% 
  mutate(size = shts*ht, size_next = shts_next*ht_next, log_size = log(size), log_size_next = log(size_next))
```

Seedlings are likely to have different responses than post-seedlings, so I'll split the data that way. To start with a relatively simple model, I'll only use continuous forest for now.

```{r}
demog_post <- 
  demog %>% 
  filter(code_notes != "sdlg (1)" | is.na(code_notes))

demog_post_cf <-
  demog_post %>% 
  filter(habitat == "CF")

demog_post_1ha <-
  demog_post %>% 
  filter(habitat == "1-ha")
```

# Function
Likelihood ratio test for random effects from Tredennick et al.  Tests observed likelihood ratio between model with and without interaction against what would be expected under null.

```{r}
sy_lrt <- function(model_y, model_sy, nsim = 500) {
  ## Loop over many iterations, simulate, compute LL: generates null distribution
  lrt <- numeric(nsim)
  
  pb <- txtProgressBar(min = 1, max = nsim, style = 3)
  for(i in 1:nsim){
    gtmp <- model.frame(model_y) # make copy of gdata
    y_star <- simulate(model_y) # simulate model (null hypothesis)
    
    #which column is response (probably always the first one, but just in case)
    resp_col <- attr(attributes(gtmp)$terms, "response")
    gtmp[ , resp_col] <- as.numeric(y_star[,1]) 
    
    # Fit models with simulated data from the null hypothesis
    tmp_null <- suppressWarnings(suppressMessages(update(model_y, data = gtmp)))
    tmp_full <- suppressWarnings(suppressMessages(update(model_sy, data = gtmp)))
    
    # Calculate log likelihoods
    ll_null <- logLik(tmp_null)
    ll_full <- logLik(tmp_full)
    lrt[i] <- 2*as.numeric(ll_full - ll_null)
    setTxtProgressBar(pb, i)
  }
  
  ## Calculate upper 95th quantile as value to exceed, for one-sided test
  upper <- as.numeric(quantile(lrt, probs = 0.95))
  
  ## Calculuate observed log likelihoods
  ll_null <- logLik(model_y)
  ll_full <- logLik(model_sy)
  obs_lrt <- 2*as.numeric(ll_full - ll_null)
  
  print(glue::glue("
                   Observed LR: {round(obs_lrt, 3)}.
                   95th quantile expected under null: {round(upper, 3)}
                   Significant interaction: {obs_lrt > upper}"))
  invisible(list(sim_lrt = lrt, q95 = upper, obs_lrt))
}
```


# Analysis (CF)

## Growth

```{r}
growth_null <- lmer(log_size_next ~ log_size + (1|plot), data = demog_post_cf)
growth_y <- lmer(log_size_next ~ log_size + (1|plot) + (1|year), data = demog_post_cf)
growth_sy <- lmer(log_size_next ~ log_size + (1|plot) + (log_size|year), data = demog_post_cf)
AICtab(growth_null, growth_y, growth_sy)
```
Significant interaction.

```{r}
#This takes a very long time and gives results qualitatively similar to comparing AIC
lrt_growth <- sy_lrt(growth_y, growth_sy, nsim = 500)
```


```{r}
ecdf(lrt_growth$sim_lrt) -> growth_dist
2*(1-growth_dist(45.689))
```


## Survival

```{r}
surv_null <- glmer(surv ~ log_size + (1|plot), family = binomial, data = demog_post_cf)
surv_y <- glmer(surv ~ log_size + (1|plot) + (1|year), family = binomial, data = demog_post_cf)
surv_sy <- glmer(surv ~ log_size + (1|plot) + (log_size|year), family = binomial, data = demog_post_cf)
AICtab(surv_null, surv_y, surv_sy)
```
no interaction

```{r}
lrt_surv <- sy_lrt(surv_y, surv_sy, nsim = 500)
```


```{r}
ecdf(lrt_surv$sim_lrt) -> surv_dist
2*(1-surv_dist(29.181))
```

## Flowering

```{r}
flwr_null <- glmer(flwr ~ log_size + (1|plot), family = binomial, data = demog_post_cf)
flwr_y <- glmer(flwr ~ log_size + (1|plot) + (1|year), family = binomial, data = demog_post_cf)
flwr_sy <- glmer(flwr ~ log_size + (1|plot) + (log_size|year), family = binomial, data = demog_post_cf)
AICtab(flwr_null, flwr_y, flwr_sy)
```
No interaction

```{r}
lrt_flwr <- sy_lrt(flwr_y, flwr_sy, nsim = 500)
```


```{r}
ecdf(lrt_flwr$sim_lrt)->flwr_dist
2*(1-flwr_dist(1.77))
```


# Analysis (1-ha)

## Growth

```{r}
growth_null_1ha <- lmer(log_size_next ~ log_size + (1|plot), data = demog_post_1ha)
growth_y_1ha <- lmer(log_size_next ~ log_size + (1|plot) + (1|year), data = demog_post_1ha)
growth_sy_1ha <- lmer(log_size_next ~ log_size + (1|plot) + (log_size|year), data = demog_post_1ha)
AICtab(growth_null_1ha, growth_y_1ha, growth_sy_1ha)
```
significant interaction

```{r}
lrt_growth_1ha <- sy_lrt(growth_y_1ha, growth_sy_1ha, nsim = 500)
```


## Survival

```{r}
surv_null_1ha <- glmer(surv ~ log_size + (1|plot), family = binomial, data = demog_post_1ha)
surv_y_1ha <- glmer(surv ~ log_size + (1|plot) + (1|year), family = binomial, data = demog_post_1ha)
surv_sy_1ha <- glmer(surv ~ log_size + (1|plot) + (log_size|year), family = binomial, data = demog_post_1ha)
AICtab(surv_null_1ha, surv_y_1ha, surv_sy_1ha)
```
No interaction or effect of year.

```{r}
lrt_surv_1ha <- sy_lrt(surv_y_1ha, surv_sy_1ha, nsim = 500)
```



## Flowering

```{r}
flwr_null_1ha <- glmer(flwr ~ log_size + (1|plot), family = binomial, data = demog_post_1ha)
flwr_y_1ha <- glmer(flwr ~ log_size + (1|plot) + (1|year), family = binomial, data = demog_post_1ha)
flwr_sy_1ha <- glmer(flwr ~ log_size + (1|plot) + (log_size|year), family = binomial, data = demog_post_1ha)
AICtab(flwr_null_1ha, flwr_y_1ha, flwr_sy_1ha)
```
significant interaction?

```{r}
lrt_flwr_1ha <- sy_lrt(flwr_y_1ha, flwr_sy_1ha, nsim = 500)
```
