---
title: "Density dependence"
author: "Eric R. Scott"
date: "2020-09-13"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(lme4)
library(bbmle)
library(car)
library(broom)
library(broom.mixed)
library(patchwork)
library(performance)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose

Start by looking for density dependence on growth, survival, and reproduction in continuous forests using the number of plants in the same grid-cell as a metric for density.

# Load Data

```{r data, echo=TRUE}
ha <- read_rds(here("analysis", "data", "derived_data", "ha_density.rds"))
ha
```

```{r}
ha <-
  ha %>% 
  group_by(ha_id_number) %>% 
  arrange(ha_id_number, year) %>% 
  ungroup() %>% 
  #scale year
  mutate(year_post = year - min(year))
```

Filter to just continuous forest

```{r}
ha_cf <- 
  ha %>% 
  filter(habitat == "CF") %>% 
  mutate(log_shts = log(shts),
         log_shts_next = log(shts_next))
```

Or filter to just 1-ha and cf

```{r}
ha_sub <-
  ha %>% 
  filter(habitat %in% c("CF", "1-ha")) %>% 
  mutate(log_shts = log(shts),
         log_shts_next = log(shts_next),
         habitat = as.factor(habitat))
```


I don't know what a height of 0 means...
Height of 0 probably means it was knocked over, or very small, or dead.

# General model structure

For each vital rate I'll include the number of shoots and habitat as fixed effects and plant ID and year as random effects.  I'm not including plot since there are few plots in each habitat type, and preliminary modeling shows the variance for plot is the lowest of those three random effects.

```{r}
ha_sub %>% 
  group_by(habitat) %>% 
  summarize(length(unique(plot)))
```


# Growth

For this I'm going to use shoot number. Very closely related to height, but better behaved distribution and not sensitive to plants being knocked over.

## Distribution

Number of shoots is discrete, but not *really* count data.  I'll try a poisson distribution and a lognormal distribution.

```{r}
mpois <- glmer(shts_next ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year_post),
                family = poisson(link = "log"),
                data = ha_sub)

check_overdispersion(mpois2) #actually underdispersed

mnorm <- glmer(shts_next ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year_post),
                family = gaussian(link = "log"),
                data = ha_sub)
```

But how bad is this convergence failure?

Is the max|grad| calculation correct?

```{r}
derivs1 <- mnorm@optinfo$derivs
sc_grad1 <- with(derivs1, solve(Hessian, gradient))
max(abs(sc_grad1))
```
No, it's actually a lot smaller than `tol`, so nothing to worry about


```{r}
AICtab(mpois, mnorm)
compare_performance(mpois, mnorm, rank = TRUE)
```

Lognormal distribution fits best according to AIC, although poisson has lower RMSE.  I guess despite being discrete, this isn't really "count" data *per se*.  The lognormal model doesn't have problems with singularity that the poisson model has.  Both have problems with heteroscedasticity.  I'll go with log-normal.

```{r}
m_growth <- mnorm
```

```{r}
x<-check_heteroscedasticity(m_growth)
plot(x)
```

## Random effects

```{r}
m_growth2 <- glmer(shts_next ~ log_shts + habitat + density_0_10 + density_10_20 + (1|year_post),
               family = gaussian(link = "log"),
               data = ha_sub)
AICtab(m_growth, m_growth2)
```

Can't remove plant ID

```{r}
m_growth3 <- glmer(shts_next ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number),
               family = gaussian(link = "log"),
               data = ha_sub)
AICtab(m_growth, m_growth2, m_growth3)
```

## Does density matter beyond immediate surroundings?

```{r}
Anova(m_growth)
```

Odd, density further out is significant, but density close is not.

```{r}
fixef(m_growth) %>% exp()
```

Both weakly positive density dependence

# Survival

Pre-reproductive census, so plants live/die in same year as `shts` and `ht`

```{r}
m_surv <- glmer(surv ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year), 
                family = binomial, data = ha_sub)
```

## Random effects

```{r}
m_surv2 <- glmer(surv ~ log_shts + habitat + density_0_10 + density_10_20 + (1|year), 
                family = binomial, data = ha_sub)
AICtab(m_surv, m_surv2)
```

Close to equivalent, so remove plant ID.  This makes sense, as plants only die once!

```{r}
m_surv3 <- glm(surv ~ log_shts + habitat + density_0_10 + density_10_20, 
                family = binomial, data = ha_sub)
AICtab(m_surv2, m_surv3)
```
Can't remove year.

## Density dependence?

```{r}
Anova(m_surv2)
fixef(m_surv2) %>% plogis()
```

Marginal, and only immediate grid cell.  Density has positive affect on survival.  So, probably just good patches.

# Flowering probability

```{r}
ggplot(ha_sub, aes(infl)) + geom_histogram()
```
Make a binary column

```{r}
ha_sub <- 
  ha_sub %>% 
  mutate(flwr = as.integer(infl>0), .before = infl)

#the way suvival is currently coded, a plant can flower and then not survive in the same year, so no need to filter out dead plants.
ha_sub %>% filter(surv == 0 & flwr == 1)
```

```{r}
m_flwr <- glmer(flwr ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year),
                family = binomial, data = ha_sub)
```

## Random effects

```{r}
m_flwr2 <- glmer(flwr ~ log_shts + habitat + density_0_10 + density_10_20 + (1|year),
                 family = binomial, data = ha_sub)
AICtab(m_flwr, m_flwr2)
```
Can't remove plant ID.


```{r}
m_flwr3 <- glmer(flwr ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number),
                 family = binomial, data = ha_sub)
AICtab(m_flwr, m_flwr2, m_flwr3)
```

Can't remove year

## Density dependence?

```{r}
Anova(m_flwr)
fixef(m_flwr) %>% plogis()
```

Ok, there's an effect, but oddly only past 20m.  Again, positive density dependence.

# TODO: 

- Check number of inflorescencesa given flowered?
- ~~check that survival is coded right so that it is a function of `ht_prev`~~
- try negative binomial distribution for shoot number
- ~~Density should be total shoot number~~
- ~~Should actually be density in previous year~~
- ~~Shoots should be poisson distribution, not log-transformed?~~
- ~~incorporate 1-ha fragments and use habitat type as a predictor.~~
- Implement FLM approach to density dependence.
